Dissecting the Role of Long-Range Amino Acid Interactions and Local Packing Density in Protein Sequence Evolution

### 2017-01-19 manuscript/voronoi/trash/mnv.tex
final article source files added

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi$^{1,2}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}

    What are the best structural determinants of protein sequence evolution? A number of site-specific structural characteristics have been proposed over the past decade to answer this question. Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability in proteins. The most commonly used measures of local packing density such as the Contact Number and the Weighted Contact Number represent by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability. Here we propose a methodology to segregate and quantify the extent to which long-range interactions determine the general patterns of sequence variability in proteins, independently of local packing density. We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein. Using a dataset of 209 monomeric enzymes, we show that the long-range amino acid interactions can explain $\sim000\%$ of variability observed in protein sequence, whereas only $\sim000\%$ of sequence variability can be attributed to site-specific packing density.

@@

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

@@

    It is notable that \citet{halle_flexibility_2002} laid a theoretical foundation for the existence of a inverse relationship between the site-specific flexibility -- often represented by the $C_\alpha$ atomic B factors -- and the local packing density in proteins. Such an inverse relationship could be intuitively understood knowing that the inverse of local packing density can serve as an estimate of the accessible free volume to each site in the protein's 3-dimensional structure, and thus a proxy measure of site-specific flexibility in protein. A crude measure of local packing density for each site can be obtained through a quantity widely known as Contact Number \citep[e.g.,][]{liao_protein_2005}. In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around the site \citep[e.g.,][]{franzosa_structural_2009},

@@

    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable parameter: the exponent of the power-law kernel ($\alpha$). The value of the exponent that results in the best predictions of B factors appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.  A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability. The specific value $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, independently of the local packing density. Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009, shahmoradi_predicting_2014}, aside from local packing density. In this regard, the two commonly-used definitions of packing density (CN \& WCN) may not be considered {\it local}. An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, independently of each other. \\

    Motivated by the existing gaps in the current understanding of the role of flexibility, packing density and long-range interactions on sequence-structure relations in proteins, here we derive a new set of site-specific structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions. In particular, we propose a methodology to derive a minimally-biased measure of local packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.  This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.

@@

\section{Protein Dataset and Site-Specific Structure$/$Sequence Variability Measures}
\label{sec:methods}

@@

    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}. (XX a comparison of CN cutoff distance with average amino acid neighbor distance is needed here to quantify evidence for the following sentence. An additional figure may also be necessary plotting cutoff distance vs. average NN distance for each protein XX). Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations. The significance and impact of these non-local interactions on sequence evolution cannot be deciphered solely from the two common definitions of packing density: CN \& WCN. Therefore, we present, in the following section, a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.

@@

    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility in Proteins}

@@

    The Voronoi cell characteristics are also strongly associated with each other. Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for. The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (c.f., Figure \ref{fig:voronoi_ER_cond}). Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to represent almost the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. However, these cell characteristics are discrete (integer) quantities and in general have limited ranges. \\

@@

    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Evolution}
    \label{sec:long_range}
    Figure \ref{fig:best_predictorER} compares the prediction power of each of the site-specific structural quantities about sequence evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes. Not shown here for brevity, similar results are also obtained for sequence entropy as the measure of sequence variability. For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains. Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability \footnote{The pairwise t-tests for all sets of correlations are available in the repository of the work (c.f., Section \ref{sec:methods})}. The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn. \ref{eqn:wcn_pwrl} also takes into account the potential long-range interactions among amino acids in different regions of protein. \\

@@

    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most widely studied structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes. Examples of important sequence--structure relations include the correlations of measures of evolutionary rates such as {\it r4sJC} used in this work and sequence entropy, with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA). In addition, we have derived a new set of site-specific characteristics from Voronoi Tessellation of protein 3D structure, some of which are capable of explaining sequence variability equally well or better than several structural quantities previously considered in the literature, such as B factor, RSA, and the traditional definitions of contact number and the weighted contact number (WCN) using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}).

    We also argued that the commonly used measures of local packing density, most importantly the Contact Number (CN) and the Weighted Contact Number (WCN), involve free adjustable parameters in the definitions which can be fine-tuned for each individual protein to obtain the best correlations between CN$/$WCN and sequence evolutionary rates. Although the best parameter values for CN \& WCN vary from one protein structure to another, the median values over the entire dataset indicate a non-negligible selection pressure due to long-range amino acid interactions on sequence evolution, independently of the local packing density.

@@

\section{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
\label{app:best_bf}

    For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain results in the best correlations with other structural and sequence properties. Specifically, the average side chain B factor outperforms the commonly used $C_\alpha$ B factor in predicting RSA, \ddg rate, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}). \\

    The observed improvements in correlations of average side-chain B factor (vs. C$_\alpha$ B factor) with other structural properties also merit further attention. It was discussed in Section \ref{sec:voro} and depicted in the plots of Figure \ref{fig:best_bf} that in general, as one moves from the B factors of atoms in the backbone of amino acid to the B factor of side-chain atoms, the correlations of B factor with other site-specific structural and sequence properties improve. In particular, the use of average side-chain B factor turned out to result in the highest correlation strengths with other site-specific properties, implying that this average B factor is likely the best representation of the overall amino acid fluctuations and flexibility in a given site in protein.  The definition of B factor and its derivation from Debye-Waller factor has been already discussed in Chapter \ref{chap:sschar}, Eqns. \ref{eqn:dwf_def}--\ref{eqn:bf_def}. \\

    The mean-square-displacement $\langle u^2 \rangle$ in Eqn. \ref{eqn:bf_def} can be decomposed into four contributing components \cite{frauenfelder_temperature-dependent_1979},

    \begin{equation}
        \label{eqn:bf_comp}
        \langle u^2 \rangle = \langle u^2 \rangle_c + \langle u^2 \rangle_d + \langle u^2 \rangle_{ld} + \langle u^2 \rangle_v,
    \end {equation}

    \noindent in which subscripts $c, d, ld, v$ refer to fluctuations due to conformational substates, diffusion, lattice disorder, and thermal vibrations respectively. The second term $\langle u^2 \rangle_d$ is generally negligible and can be ignored in Eqn. \ref{eqn:bf_comp}. Of particular interest to this study is the first term, which is also typically the major contributor to the overall value of the atomic B factor, specially in high-resolution X-ray crystallography of proteins. This term represents the positional displacements of the atom of interest together with other atoms in the amino acid between many different conformational substates of the protein, with the transition probability between the substates governed by the Boltzmann distribution. Compared to atomic coordinates, there are comparatively fewer restraints on the atomic B factors during X-ray crystallography refinement process, and thus in this regard B factor is generally considered as the {\it error sinks} for static and dynamic disorder and various kinds of model errors in the refinement process \cite{read_structure-factor_1990}. The noise and model uncertainty contributions to the atomic B factors in particular increase with decreasing the resolution of the X-ray crystallography. Better resolution in general corresponds to lower average B factors for the entire structure of the protein \cite{read_structure-factor_1990}. \\

    Although the extraction of conformational fluctuations from noise in B factors seems a daunting task \cite{read_structure-factor_1990}, the effects of noise, model error and uncertainties due to limited X-ray crystallography resolution can be minimized by averaging B factors over the entire amino acid in a given site: To expand on this, consider the contribution of conformational fluctuations between different substates to be approximately the same for all atoms in the amino acid. The conformational fluctuations can be regarded as the collective motion of all atoms in the amino acid, on top of which there are noise fluctuations in each of the atoms. These collective motions are the type of fluctuations in B factors that are expected to reflect the biologically relevant and important factors for the proper functioning of the protein. The stochastic noise in the fluctuations is often assumed to have an isotropic Gaussian origin. Therefore, averaging over the atomic B factors in each individual amino acid essentially results in higher Signal-to-Noise Ratio (SNR) in the measurement of the amino acid conformational fluctuations. Figure \ref{fig:bfC_bfAA_ratio} illustrates how this averaging over all atomic B factors increases the SNR in measuring the fluctuations due to conformational substate transitions of the amino acid. \\

    % Figure goes here

    To expand further on this, a simple argument may be given to explain the observed strongly-positive approximately-linear correlation between the two parameters in the plot of Figure \ref{fig:bfC_bfAA_ratio}. The contributions to the atomic B factor values of the $i^{th}$ atom in the amino acid in the $j^{th}$ site in a given protein can be assumed to originate from two major sources: conformational substates and stochastic noise due to model uncertainties in refinement process and limited resolution of the X-ray crystallography,

    \begin{equation}
        \label{eqn:bf_comp}
        \langle u^2 \rangle_{ij} = \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij}.
    \end {equation}

    For simplicity and without loss of generality, one can assume that the contribution of fluctuations due to conformational substate transitions is approximately the same for all atomic B factors in a given amino acid residing the $j^{th}$ site. In other words, the term $\langle u^2 \rangle_{\text{substates},ij}$ in the above equation has almost the same value $\langle u^2 \rangle_{\text{substates},j}$ for all atoms in the amino acid in the $j^{th}$ site in protein. Thus, the average B factor for the entire amino acid molecule of size $N_j$ atoms would be,

    \begin{eqnarray}
        \label{eqn:bf_avg}
        \langle u^2 \rangle_{j} &=& \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{substates},ij} + \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \langle u^2 \rangle_{\text{noise},ij} \nonumber \\
                                &=& \langle u^2 \rangle_{\text{substates},j} + \frac{1}{N_j} \sum_{i=1}^{N_j} \mu_{\text{noise},j}
    \end {eqnarray}

    \noindent in which $\mu_{\text{noise},j}$ is the average noise in the $j^{th}$ amino acid. The ratio of the B factor of the $ij^{th}$ atom to the average B factor of the $j^{th}$ site in protein can be approximated as,

    \begin{eqnarray}
        \label{eqn:bf_ratio}
        \frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} &\simeq& \frac{\langle u^2 \rangle_{substates,j} + \langle u^2 \rangle_{noise,ij}}{\langle u^2 \rangle_{substates,j} + \mu_{noise,j}} \\
                                 &=& \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \nonumber \\
                                 &+& \bigg(\frac{\langle u^2 \rangle_{\text{noise},ij}}{\langle u^2 \rangle_{\text{substates},j}} \bigg)
                                 \frac{1}{1 + \mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j}} \label{eqn:bf_ratio1} \\
                                 &\simeq& 1 - \frac{\mu_{noise,j}}{\langle u^2 \rangle_{\text{substates},j}}, \label{eqn:bf_ratio2}
    \end {eqnarray}

    \noindent where from line \ref{eqn:bf_ratio1} to \ref{eqn:bf_ratio2}, an assumption was made that the second term in line \ref{eqn:bf_ratio1} could be neglected compared to the first term and that the noise compared to conformational fluctuation is small, that is, $\mu_{noise,j} / \langle u^2 \rangle_{\text{substates},j} < 1$ (an error of $0.2\AA$ corresponds approximately to $1\AA$ increase in B factor \cite{read_structure-factor_1990}). Knowing that the average noise across different amino acids is approximately the same \cite{frauenfelder_temperature-dependent_1979}, that is $\mu_{noise,j}\sim \mu_{noise}$, and that the noise due to X-ray crystallography almost negatively linearly correlates with crystallography resolution in the range $\sim1-3~[\AA]$ \cite{read_structure-factor_1990}, that is $\mu_{noise}\propto -\text{resolution}$, a positive approximately-linear relationship between the average of the B factor ratios over the entire amino acids in the protein structure and the X-ray crystallography resolution would be obtained,

    \begin{eqnarray}
        \frac{\text{BF}_C}{\text{BF}_{AA}} &=& \frac{1}{L}\sum_{j=1}^{L}\frac{\langle u^2 \rangle_{ij}}{\langle u^2 \rangle_{j}} \label{eqn:avg_bf_ratio} \\
                                           &\propto& -\mu_{noise} \sum_{j=1}^{L}\frac{1}{\langle u^2 \rangle_{substates,j}} \label{eqn:avg_bf_ratio1} \\
                                           &\propto& \text{resolution}
    \end {eqnarray}

    \noindent in which $L$ represents the length of the protein sequence. The summation term in line \ref{eqn:avg_bf_ratio1} would not influence this linear relationship, causing only scatter in the relation, so long as the length of the protein not does impose limitations on the resolution of X-ray crystallography of proteins. In general, however this may not be the case. For the sample of 209 proteins considered here, there exists indeed a weak Spearman's correlation coefficient of $\rho\sim0.2$ between protein length ($L$) and resolution. Figure \ref{fig:bfC_bfAA_ratio} illustrates the relationship between the average B factors ratio and the resolution in the dataset, using atom $C$ in the backbone of all amino acids in proteins representing the $i^{th}$ atom in the notation of Eqn. \ref{eqn:avg_bf_ratio}. It is also notable that the the atomic fluctuations due to conformational substates may not be exactly the same for all atoms in an amino acid in a given site in protein. Indeed, one may expect the conformational fluctuations in the backbone atoms would be less significant compared to conformational fluctuations of side-chain atoms. \\

    Although averaging B factor over the entire amino acid atoms would reduce the noise further than averaging over side-chain atoms, the functionally important conformational fluctuations that are better captured by the side-chain atomic B factors would compensate for the increase in the noise, such that overall, the B factors averaged over side-chain atoms results in slightly better correlations with sequence variability and other relevant structural characteristics depicted in Figure \ref{fig:best_bf}. \\

@@

    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells. Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties. These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$). In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains. Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together. In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid. The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations. The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}. Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all. \\

@@

    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}). \\

    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells. Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}). \\

@@

>>>>

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Dissecting the Role of Long-Range Amino Acid Interactions and Local Packing Density in Protein Sequence Evolution}
\vfill
\Author{Amir Shahmoradi$^{1,2}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}

    What are the best structural determinants of protein sequence evolution? A number of site-specific structural characteristics have been proposed over the past decade to answer this question. Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability in proteins. The most commonly used measures of local packing density such as the Contact Number and the Weighted Contact Number represent by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability. Here we propose a methodology to segregate and quantify the extent to which long-range interactions determine the general patterns of sequence variability in proteins, independently of local packing density. We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein. Using a dataset of 209 monomeric enzymes, we show that the long-range amino acid interactions can explain $\sim10\%$ of variability observed in protein sequence, whereas $\sim30\%$ of sequence variability can be attributed to site-specific packing density.

@@

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

@@

    A crude measure of site-specific packing density for each site in protein is often obtained through a quantity widely known as Contact Number \citep[e.g.,][]{liao_protein_2005}. Alternative packing density measures have also been proposed leading to better predictions of sequence evolution and site-specific flexibility, a prime example of which is the Weighted Contact Number (WCN) proposed by \citet{lin_deriving_2008}. These proxy measures of packing density however, involve adjustable free parameters in their definitions. Moreover, the packing density as defined by these quantities is more representative of the local environment around the site of interest, and does necessarily correspond to the site-specific packing density. \\

    Motivated by the existing gaps in the current understanding of the role of site-specific packing density and long-range amino acid interactions in protein sequence evolution, here we derive a new set of structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions. In particular, we propose a methodology to derive a minimally-biased measure of local packing density based upon which the role of long-range amino-acid interactions on sequence variability can be segregated and isolated from the effects of site-specific packing density.  This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility. In this regard, the two commonly-used definitions of local packing density (CN \& WCN) may not be considered {\it site-specific}. An important question therefore remains unanswered as to what extent the site-specific packing density and long-range interactions influence sequence variability, independently of each other.  \\

\section{Methods}
\label{sec:methods}

    \subsection*{Local Packing Density Definition and Measurement}

    The local packing density is commonly measured by a quantity known as Contact Number (CN). In its simplest mathematical form, the Contact Number for a given site $i$ in a protein of amino acid sequence length of $N$ is defined as the number of amino acids within a fixed radius $r_0$ of neighborhood around the site \citep[e.g.,][]{franzosa_structural_2009},

@@

    Although WCN is in general a better predictor of site-specific sequence variability, the proposed definition of WCN still involves an adjustable parameter: the exponent of the power-law kernel ($\alpha$). The value of the exponent that results in the best predictions appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.  A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability. The specific value $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, independently of the local packing density. Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on sequence variability \citep[e.g.,][]{franzosa_structural_2009, shahmoradi_predicting_2014}, aside from site-specific packing density. \\

@@

    \subsection*{Protein Dataset and Site-Specific Structure$/$Sequence Variability Measures}
    \label{sec:methods}

@@

    The results for the dataset of $209$ monomeric enzymes are plotted in Figure \ref{fig:cnwcnp} and the values of the free parameters of CN and WCN that yield the strongest correlations are tabulated in Table \ref{tab:best_params}. Evidently, the best-performing free parameters of both models indicate a non-negligible contribution of the long-range amino acid interactions, beyond the immediate neighborhood of the site, to the strengths of the observed correlations. The significance and impact of these non-local interactions on sequence evolution cannot be deciphered solely from the two common definitions of packing density: CN \& WCN. Therefore, we present, in the following section, a methodology to segregate the role of local packing density from long-range interactions and quantify their individual contributions to site-specific sequence variability measures.

@@

    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility}

@@

    The Voronoi cell characteristics are also strongly associated with each other. Although the cell volume and area are almost identically the best correlating variables with ER, the cell volume does not exhibit any significant independent correlation with ER once the cell area is controlled for. The median strength of the partial correlation of volume with ER, while controlling for area is centered at $\sim0.0$ (Figure \ref{fig:voronoi_ER_cond}). Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to represent almost the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. However, these cell characteristics are discrete (integer) quantities and in general have limited ranges. \\

@@

    \subsection*{Effect of Long-Range Amino Acid Interactions on Sequence Evolution}
    \label{sec:long_range}
    Figure \ref{fig:best_predictorER} compares the prediction power of each of the site-specific structural quantities about sequence evolutionary rates ({\it r4sJC}) for the dataset of $209$ monomeric enzymes. Not shown here for brevity, similar results are also obtained for sequence entropy as the measure of sequence variability. For comparison, the results for WCN calculated using the C$_\alpha$ atomic coordinates are also illustrated in the plot, in addition to WCN calculated from the coordinates of the geometric centers of side chains. Notably, the quantity WCN outperforms all other structural quantities in explaining site-specific sequence variability. In particular, a pairwise t-test between the correlation strengths of r4sJC-WCN and the correlation of r4sJC with the inverse of Voronoi cell volume yields a p-value of $<10^{-16}$. The better performance of WCN compared to local packing density as measured by the inverse of Voronoi cell volume may not be surprising, knowing that WCN by its definition in Eqn. \ref{eqn:wcn_pwrl} also takes into account the potential long-range interactions among amino acids in different regions of protein. \\

@@

    Throughout the previous sections, we carried out a comprehensive analysis and comparison of the most widely studied structural determinants of sequence variability, using a dataset of $209$ monomeric enzymes. Examples of important sequence--structure relations include the correlations of measures of sequence variability -- such as evolutionary rates (e.g., {\it r4sJC}) and sequence entropy -- with measures of residue Contact Number (CN \& WCN) and Relative Solvent Accessibility (RSA). In addition, we derived a new set of site-specific characteristics from the Voronoi partitioning of protein's 3D structure, some of which are capable of explaining sequence variability equally well or better than several structural quantities that were previously considered in the literature -- including B factor, RSA, and the traditional definitions of CN and WCN using $C_\alpha$ atomic coordinates (Figure \ref{fig:best_predictorER}). \\

    The commonly used measures of local packing density, most importantly CN and WCN, involve free adjustable parameters in their definitions that can be fine-tuned for each individual protein to obtain the strongest correlations between CN$/$WCN and sequence variability.   Although the best parameter values for CN \& WCN vary from one protein structure to another, the median of the values over the entire dataset clearly indicate a non-negligible selection pressure on protein sequence solely due to long-range amino acid interactions, independently of the local packing density. Unlike CN and WCN, the site-specific packing density as defined by the inverse of Voronoi cell volume in this work, represents only the contributions from the nearest neighbors of each site in protein, that is, the first coordination shell.  \\

    Although the site-specific packing density as defined by the inverse of the Voronoi cell volume excludes the effects of long-range interactions beyond the first coordinations shell, it remains as best predictor of sequence variability in our dataset among all other structural determinants, most importantly RSA. Contrary to the findings of this work however, some other studies \citep[e.g., ][]{franzosa_structural_2009, scherrer_modeling_2012, shahmoradi_predicting_2014} have argued for RSA as the main determinant of sequence evolution with local packing density having a peripheral role. Our conclusion is that RSA and the local packing density -- once corrected for long-range interactions -- in principle represent the same characteristics of the local environment in protein, that is, both quantities are proxy measures of the number of neighboring amino acids in the first coordination shell. This argument is also further supported by running a pairwise t-test between the correlation strengths of RSA with evolutionary rates and the correlation strengths of the inverse Voronoi cell volume with evolutionary rates for the entire protein dataset, with a resulting p-value $\sim0.29$. An important question however remains unanswered in this work and merits further study, as to what extent the findings of the presented analysis -- based on a dataset of 209 monomeric enzymes -- can be generalized to other types of proteins in nature.

@@

@@

    Here in this work, all possible choices of the representative set of atomic coordinates were considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells. Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties. These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$). In addition, representative coordinates for each site in protein can be also calculated by averaging over the coordinates of all heavy atoms in the side chains. Also calculated was a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together. In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid. The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate as defined by \citet{echave_relationship_2014}, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and the average backbone Hydrogen bond energy with different measures of WCN and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn} and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations. The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}. Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all. The Hydrogen bonds were identified and the corresponding energies were calculated according to the prescription of \citet{kabsch_dictionary_1983}. \\

@@

    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible sets of coordinates, the average over coordinates of all heavy atoms of each individual side-chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}). \\

    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells. For brevity, only the results for the cell area are given here. Specifically, the cell area from average side-chain coordinates outperforms cell area calculated based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}). \\

@@

%\section{Average Side-Chain B Factors as the Best Representation of Local Fluctuations of Amino Acids in Proteins}
%\label{app:best_bf}
%
